Supporting Information

### 2015-08-21 review1.Rmd
edits to response letter

In this manuscript, Duursma et al. present findings on the response of LAI and litter production to elevated atmospheric CO2 in EucFACE, a free air CO2 enrichment experiment established in a Eucalyptus woodland. The authors had expected elevated CO2 to increase LAI and litter production. In contrast to these expectations, the authors found CO2 to have no significant effect on either LAI or litter production over the 28-month experiment. However, they did find that CO2 extended leaf lifespan by approximately two months.

@@

**We now use pre-treatment BA as a covariate in a mixed-effects model. We found no CO2 or CO2 x time effect on LAI when BA was used as a covariate. There is also no effect of CO2 when BA is not included as a covariate, but due to the tight relationship between BA and LAI, using BA as a covariate increases the ability to find a CO2 effect. On this point, we have also added a new figure in the supporting information (Fig. S5) that addresses the minimum detectable difference with a new method. Fig.2 now shows the LAI timeseries instead of the anomaly as before, and we have amended the text in various places to account for these changes.**

The only statistically significant response of LAI that would be expected under the current statistical design would be from a CO2 x time interaction. While no statistics are reported, an interaction may have been detected but attributed exclusively to an insect outbreak, whereas it also has the potential to due to a real response. However, the inclusion of the post-outbreak data makes this difficult to discern. The outbreak confounds the effect of CO2 with that of the outbreak, and thus data from the infested plot should only be considered up until the outbreak occurred.

**As noted above, we have redone the analysis and find no CO2 or CO2 x time effect. Although there was a brief period when the 95% confidence intervals were not overlapping (for the anomaly in the previous manuscript, though not when LAI is considered, see new Fig.2), this brief period did not result in a significant overall CO2 x time effect. For this reason we have not removed the insect attack from the study (it would have been appropriate if we did find a CO2 x time effect), since it does not confound the overall findings.**

The techniques used to collect litter are also concerning. The authors deployed eight 0.2 m2 litter traps in each ~123 m2 plot and used the litter collected from these traps to estimate litter production. However, considering the size of each individual trap and scale of the experiment, there is significant doubt that the traps provided a representative sample of each plot. Additionally, the analyses performed using litter data make several assumptions that are unlikely to hold.

@@

**The size, number, and relative coverage of the litterfall traps at EucFACE is consistent with the design of other forest FACE studies (see Table below), with in fact the largest trap size compared to other sites. The size and coverage of the traps is therefore fully justified. The litter traps were placed in somewhat random locations across the ring, but never within 2 meters of the plot edge, to minimize contamination. We, as in all other FACE sites, cannot be completely certain that contamination did not occur, but it is likely that the vast majority of litter in each basket was produced in that FACE ring.**

@@

Another issue lies in the inclusion of an add-on analysis comparing two LAI estimation techniques. While useful, this comparison should be its own methods paper and does not align well with the scope of the rest of the manuscript. Currently, it adds to an already lengthy paper and distracts from the authors’ main points.

**We have moved this analysis, including the methods, in its entirety to the Supporting Information. We believe this is interesting information for others concerned with LAI estimation, but agree it is a secondary objective only. We only briefly mention in the Methods that the two methods gave comparable results.**

Finally, the manuscript requires extensive editing for structure and flow.

@@

The introduction does not sufficiently establish the expectation of litter responses.

@@

The methods are incomplete and unclear. Specifically, there are many portions of the Results section that should be included in Materials and Methods.

@@

Additionally, it is unclear if the authors’ used repeated measures to analyze their data, which would be necessary considering the experimental design.

@@

Beyond these issues, the manuscript also should be revised for clarity and brevity.

@@

Specific Comments (line number given):
  Abstract
l.26 What do you mean by (hot summer), and is it necessary? This is not how you describe the climate in the methods.

@@

**Changed to LAI throughout the manuscript.**

@@

**FACE was defined in the 2nd paragraph of the definition, here the definition of EucFACE is hopefully obvious from the statement that the woodland is Eucalyptus-dominated.**

@@

**Added '(unmanaged)'.**

@@

**I HAVE NOT DONE THIS - CAN SOMEONE TELL ME WHAT IS DIFFICULT ABOUT IT???**

@@

**We have responded to the data analysis concerns above.**

@@

**We disagree that the outbreak confounded the results; we did not find effects of CO2 on LAI despite the outbreak. Although we did not set out to examine the outbreak, it occurred nonetheless and we think it is informative as the outbreak reflects natural dynamics of Eucalyptus-dominated woodlands. Also, we do not discuss the outbreak in any sort of detail, so removing it from the figure does not shorten the manuscript.**

@@

**We have responded to the concerns about data analysis above.**

@@

**Moved to methods (and edited for clarity). We disagree that this method 'assumes litter is grown and lost in the same time period'. Basically, gross leaf production must be equal to the net change in LAI (=the pool of leaves in the canopy) plus the amount of shedded leaf area, as measured by litter production. And to  "this calculation could indicate that a period of reduced Lt and large litter fall was associated with positive leaf production", this is only true if the reduction in LAI is less than the litter produced. In that case the canopy must have produced new leaves in that period to complete the sum of dLAI/dt = leafgrowth - litter. There are no assumptions on when the leaves were produced that show up in the litter traps at that time. We now explain this mass balance consideration in the Methods.**

@@

**The previous analysis was essentially equivalent to using BA as a covariate, however this is now irrelevant since we have redone the analysis.**

@@

**It was not the purpose of the manuscript to generalize to other FACE sites, but we feel it is relevant to compare the results of this study to the other FACE sites. This is informative because the results from the EucFACE do not fit well with a generalization across other FACE sites as presented by Norby and Zak (2011).**

>>>>

*Reviewer comment 1* : In this manuscript, Duursma et al. present findings on the response of LAI and litter production to elevated atmospheric CO2 in EucFACE, a free air CO2 enrichment experiment established in a Eucalyptus woodland. The authors had expected elevated CO2 to increase LAI and litter production. In contrast to these expectations, the authors found CO2 to have no significant effect on either LAI or litter production over the 28-month experiment. However, they did find that CO2 extended leaf lifespan by approximately two months.

@@

**Our response : We now use pre-treatment BA as a covariate in a mixed-effects model. We found no CO2 or CO2 x time effect on LAI when BA was used as a covariate. There is also no effect of CO2 when BA is not included as a covariate, but due to the tight relationship between BA and LAI, using BA as a covariate increases the ability to detect a CO2 effect. On this point, we have also added a new figure in the supporting information (Fig. S5) that addresses the minimum detectable difference with a new method, which is 0.05 m2 m-2 or 6%. Fig.2 now shows the LAI timeseries instead of the anomaly as before, and we have amended the text in various places to account for these changes.**

*Reviewer comment 2* : The only statistically significant response of LAI that would be expected under the current statistical design would be from a CO2 x time interaction. While no statistics are reported, an interaction may have been detected but attributed exclusively to an insect outbreak, whereas it also has the potential to due to a real response. However, the inclusion of the post-outbreak data makes this difficult to discern. The outbreak confounds the effect of CO2 with that of the outbreak, and thus data from the infested plot should only be considered up until the outbreak occurred.

**Our response : As noted above, we have redone the analysis and find no CO2 or CO2 x time effect over the entire 28-month sampling period (Fig. 2) that includes the 3 to 4 month period towards the end of the study when insect-mediated leaf loss was observed. Although there was a brief period when the 95% confidence intervals were not overlapping (for the anomaly in the previous manuscript, though not when LAI is considered, see new Fig.2), this comparatively brief period did not result in a significant overall CO2 x time effect. For this reason we have not removed the insect attack from the study (it would have been appropriate if we did find a CO2 x time effect), since it does not confound the overall findings. In any case, removal of ring 6 (one of the ambient rings) during this period has no effect on CO2 or CO2 x time effects.**

*Reviewer comment 3 and 4* : The techniques used to collect litter are also concerning. The authors deployed eight 0.2 m2 litter traps in each ~123 m2 plot and used the litter collected from these traps to estimate litter production. However, considering the size of each individual trap and scale of the experiment, there is significant doubt that the traps provided a representative sample of each plot. Additionally, the analyses performed using litter data make several assumptions that are unlikely to hold.

@@

**Our response : The size, number, and relative coverage of the litterfall traps at EucFACE is consistent with the design of other forest FACE studies (see Table below), with in fact the largest trap size compared to other sites. The size and coverage of the traps is therefore fully justified. The litter traps were placed in somewhat random locations across the ring, but never within 2 meters of the plot edge, to minimize contamination. We, as in all other FACE sites, cannot be completely certain that contamination did not occur, but it is likely that the vast majority of litter in each basket was produced in that FACE ring.**

@@

**Additionally, we provide a more detailed description of our custom-made litter traps.**

*Reviewer comment* 5 : Another issue lies in the inclusion of an add-on analysis comparing two LAI estimation techniques. While useful, this comparison should be its own methods paper and does not align well with the scope of the rest of the manuscript. Currently, it adds to an already lengthy paper and distracts from the authors’ main points.

**Our response : We have moved this analysis, including the methods, in its entirety to the Supporting Information. We believe this is interesting information for others concerned with LAI estimation, but agree it is a secondary objective only. We only briefly mention in the Methods that the two methods gave comparable results.**

*Reviewer comment* 6 :Finally, the manuscript requires extensive editing for structure and flow.

@@

*Reviewer comment* 7 :The introduction does not sufficiently establish the expectation of litter responses.

@@

*Reviewer comment* 8 :The methods are incomplete and unclear. Specifically, there are many portions of the Results section that should be included in Materials and Methods.

@@

*Reviewer comment* 9 :Additionally, it is unclear if the authors’ used repeated measures to analyze their data, which would be necessary considering the experimental design.

@@

*Reviewer comment* 10 :Beyond these issues, the manuscript also should be revised for clarity and brevity.

@@

**Specific Comments** (line number given):
  Abstract
l.26 What do you mean by (hot summer), and is it necessary? This is not how you describe the climate in the methods.

@@

**We keep the abbreviation L because it is more correct to use single letter abbreviations in equations, and it is confusing to use both L (in equations) and LAI (in the text). Although LAI is standard, many recent papers use L instead.**

@@

**We have added another definition of FACE in the Introduction.**

@@

**We have removed this sentence, as it would take too much space to give a detailed account of the site history.**

@@

**We have edited this section for clarity.**

@@

**We have responded to the data analysis concerns in comment 2.**

@@

**We disagree that the outbreak confounded the results; we did not find effects of CO2 on LAI despite the outbreak. Although we did not set out to examine the outbreak, it occurred nonetheless and we think it is informative as the outbreak reflects natural dynamics of Eucalyptus-dominated woodlands. Also, we do not discuss the outbreak in any sort of detail, so removing it from the figure does not shorten the manuscript. Furthermore, removal of ring 6 does not alter treatment effects.**

@@

**We have responded to the concerns about data analysis in comment 2**

@@

**Moved to methods (and edited for clarity). We disagree that this method 'assumes litter is grown and lost in the same time period'. Basically, gross leaf production must be equal to the net change in LAI (=the pool of leaves in the canopy) plus the amount of shed leaf area, as measured by litter production. And to  "this calculation could indicate that a period of reduced Lt and large litter fall was associated with positive leaf production", this is only true if the reduction in LAI is less than the litter produced. In that case the canopy must have produced new leaves in that period to complete the sum of dLAI/dt = leafgrowth - litter. There are no assumptions on when the leaves were produced that show up in the litter traps at that time. We now explain this mass balance consideration in the Methods.**

@@

**The previous analysis was essentially equivalent to using BA as a covariate, however this is now irrelevant since we have redone the analysis using BA as a covariate (see commen 2).**

@@

**It was not the purpose of the manuscript to generalize to other FACE sites, but we feel it is relevant to compare the results of this study to the other FACE sites. This is informative because the results from the EucFACE do not fit well with a generalization across other forest FACE sites as presented by Norby and Zak (2011).**
